---
title: "0. Setup"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file will walk through the installation of any needed packages. We will consider three types of packages:

 * **CRAN** - this is where most R packages reside, and it has extensive automated testing to ensure quality.
 * **GitHub** - these are more "bleeding edge" packages that may not have as much testing.
 * **BioConductor** - these are packages designed for computational biology applications.

General notes

 * If given the choice, do not compile packages from source (it's more error-prone).
 * If there's a corruption error message, try restarting your R session and trying again (Session -> Restart R).
 * If asked, try to update existing packages. But if that causes errors, run it again and don't update existing packages.

## Check requirements

```{r check_requirements}
min_r_version = 3.6

r_version = as.numeric(paste0(R.version["major"], ".", as.integer(R.version["minor"])))

if (r_version < min_r_version) {
  cat(paste0("Error: your R software is out of date.\n",
      "Minimum version required: ", min_r_version, "\n",
      "R version detected: ", r_version, "\n",
      "Please download and install the latest version from: https://cloud.r-project.org/"))
} else {
  cat("Your R software is up to date:", r_version)
}
```

## Package installation

### Cran packages

```{r cran_packages}
packages = list(
  cran = c("h2o",         # glrm, anomaly forests
           "dbscan",      # hdbscan
           "poLCA",       # latent class analysis
           "umap",        # umap
           "BiocManager", # needed to install hopach
           "rio",         # import / export
           "dplyr",       # data manipulation
           "ggplot2",     # plotting
           "remotes")     # installation of packages from GitHub
)

# Try to load each package and save the result.
(install_status = suppressWarnings(sapply(packages$cran, require, character.only = TRUE, quietly = TRUE)))

# Check if any packages still need to be installed, and if so, try to install them.
if (sum(!install_status) > 0) {
  cat("The following CRAN packages need to be installed:",
      paste0(paste(names(install_status)[!install_status], collapse = ", "), ".\n"))
  
  # Install any needed packages.
  install.packages(packages$cran[!install_status],
                   # Specify repository so that we can install when knitting this file.
                   repos = "http://cran.rstudio.com")
  
  # Load the newly installed packages.
  recheck = suppressWarnings(sapply(packages$cran[!install_status], require, character.only = TRUE, quietly = TRUE))
  
  if (sum(!recheck) > 0) {
    cat("Failed to install the following CRAN packages:",
        paste0(paste(names(recheck)[!recheck], collapse = ", "), "."),
        "\nPlease try installing them manually with install.packages()",
        "and closely reviewing any error messages.\n")
  } else {
    cat("CRAN packages installed successfully.\n")
  }
} else {
  cat("All CRAN packages are installed and loaded.\n")
}

```

### GitHub packages

Note that the ck37r package is now also on CRAN, but it's good to show GitHub installation for posterity.

```{r github_packages}
packages$github = c("ck37r" = "ck37/ck37r")

# Try to load each package and save the result.
install_status = suppressWarnings(sapply(names(packages$github),
                                          require, character.only = TRUE, quietly = TRUE))

# Check if any packages still need to be installed, and if so, try to install them.
if (sum(!install_status) > 0) {
  cat("The following GitHub packages need to be installed:",
      paste0(paste(names(install_status)[!install_status], collapse = ", "), "."))
  
  # Install any needed packages.
  remotes::install_github(packages$github[!install_status])
  
  # Load the newly installed packages.
  recheck = suppressWarnings(sapply(names(packages$github)[!install_status],
                                    require, character.only = TRUE, quietly = TRUE))
  
  if (sum(!recheck) > 0) {
    cat("Failed to install the following GitHub packages:",
        paste0(paste(packages$github[names(recheck)[!recheck]], collapse = ", "), "."),
        "\nPlease try installing them manually with remotes::install_github()",
        "and closely reviewing any error messages.\n")
  } else {
    cat("GitHub packages installed successfully.\n")
  }
} else {
  cat("All GitHub packages are installed and loaded.\n")
}
```

### Bioconductor packages

```{r biocon_packages}
packages$bioconductor = c("hopach")

# Try to load each package and save the result.
(install_status = suppressWarnings(sapply(packages$bioconductor,
                                          require, character.only = TRUE, quietly = TRUE)))

# Check if any packages still need to be installed, and if so, try to install them.
if (sum(!install_status) > 0) {
  cat("The following Bioconductor packages need to be installed:",
      paste0(paste(names(install_status)[!install_status], collapse = ", "), "."))
  
  # Install any needed packages.
  BiocManager::install(packages$bioconductor[!install_status])
  
  # Load the newly installed packages.
  recheck = suppressWarnings(sapply(packages$bioconductor[!install_status],
                                    require, character.only = TRUE, quietly = TRUE))
  
  if (sum(!recheck) > 0) {
    cat("Failed to install the following Bioconductor packages:",
        paste0(paste(packages$github[names(recheck)[!recheck]], collapse = ", "), "."),
        "\nPlease try installing them manually with BiocManager::install()",
        "and closely reviewing any error messages.\n")
  } else {
    cat("Bioconductor packages installed successfully.\n")
  }
} else {
  cat("All Bioconductor packages are installed and loaded.\n")
}
```

## Slides

Now open up the slides available here: [https://dlab-berkeley.github.io/Unsupervised-Learning-in-R/slides.html](https://dlab-berkeley.github.io/Unsupervised-Learning-in-R/slides.html)
