---
title: "HOPACH: hierarchical medoid clustering"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Highlights

* Hierarchical cluster algorithm that is divisive (top down), but also will consider combining clusters if beneficial.
* Based on medoids: each cluster is based on a representative observation with the smallest distance to other cluster members.
* Orders clusters and observations within clusters based on the distance metric.
* The hierarchical tree splits don't have to be binary - could split into 3 or more clusters.

## Load data

```{r load_data}
# From 1-clean-data.Rmd
data = rio::import("data/clean-data-imputed.RData")

str(data)

# Convert factors to indicators
result = ck37r::factors_to_indicators(data, verbose = TRUE)
data = result$data
```

## Basic HOPACH

```{r basic_hopach}
library(hopach)

# We use the numeric data matrix here, which has converted factors to indicators.
distances = distancematrix(data,
                           d = "euclid",
                           # d = "cosangle",
                           na.rm = TRUE)

# This is an n x n matrix.
dim(distances)

hobj = hopach(data, dmat = distances,
              d = "euclid")
              # d = "cosangle")


# Number of clusters identified.
hobj$clust$k  

# Review sizes of each cluster.
hobj$clust$sizes 

# This plot is recommended but does not seeem that useful.
# Red = small distance, white = large distance.
dplot(distances, hobj, ord = "final",
      main = "Distance matrix", showclusters = FALSE)  

# Bootstrap analysis
# TODO: identify how to set seed.
boot_result = boothopach(data, hobj, B = 100)

# We want to see large bands of the same color, which indicate cluster stability.
# May need to click "zoom" to view the plot.
bootplot(boot_result, hobj, ord = "bootp",
         main = "Bootstrap plot", showclusters = FALSE)

# Extract medoids - representative observations for each cluster.
# These are the row indices.
hobj$clustering$medoids
```

## Challenges

1. Change the distance metric to "cosangle" and re-run. How do your results compare?
2. Cluster the variables rather than the observations by transposing the data frame with `t()` and re-running the distance calculation and clustering. Examine the medoids to find the representative variables in the dataset.

## Resources

* [Bioconductor's HOPACH package](https://www.bioconductor.org/packages/release/bioc/vignettes/hopach/inst/doc/hopach.pdf)

## References

van der Laan, M. J., & Pollard, K. S. (2003). A new algorithm for hybrid hierarchical clustering with visualization and the bootstrap. Journal of Statistical Planning and Inference, 117(2), 275-303.

Pollard, K. S., & van der Laan, M. J. (2005). Cluster analysis of genomic data. In Bioinformatics and Computational Biology Solutions Using R and Bioconductor (pp. 209-228). Springer, New York, NY.
